Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-02-19 Manuscript/HI_Model_ms.Rmd
updated manuscript files

The State of Hawaiʻi exemplifies the challenges associated with projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances (Fig. S3), with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016]. Although these past trends surely inform the future impact of climate and land-use change on ecosystem carbon balance, high spatial and temporal heterogeneity complicates realistic projection efforts. To date only one study has attempted to integrate land-use, climate, and natural disturbances into future projections of Hawaiian ecosystem carbon balance, with projections limited to the mid-21st century under a single land-use change scenario and moderate radiative forcing [SRES A1B, equivalent to RCP 6; @selmants_baseline_2017].

@@

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### *2.1 Study area*
The spatial extent of this study was the terrestrial portion of the seven main Hawaiian Islands (figure 1), a total land area of 16,554 km^2^. We subdivided this landscape into a grid of 264,870 simulation cells, each of which was 250 x 250 m in size. Each simulation cell was assigned to one of 210 possible state types based on the unique combination of three moisture zones (dry, mesic, and wet; figure S1), seven islands, and ten discrete land cover classes (figure 1).

```{r fig.align = 'center', out.width = "99%", fig.cap = "Land cover classification of the seven main Hawaiian Islands, adapted from Jacobi et al (2017). Agriculture in this map combines herbaceous and woody crops, but these two crop types are treated as separate land cover classes in the simulation model. Water and Wetland land cover classes are not shown."}
knitr::include_graphics("./fig_images/fig1_HI_LC.png")
```

### *2.2 States and transitions*
We developed two land-use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transitions between state types were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, tree plantation, and shrubland state types each had multiple transition pathways, while the barren state type could only transition to developed (i.e., urbanization). There was no transition pathway out of an urbanized (developed) state. Water and wetland state types remained static throughout the simulation period.

@@

The wildfire transition sub-model was modified from @daniel_state-and-transition_2016 by incorporating a new 21-year historical wildfire spatial database of the Hawaiian Islands (figure S2). We used this new spatial database to calculate historical wildfire size distribution and ignition probabilities for each unique combination of moisture zone (figure S1), island, and state type (figure 1) for the years 1999-2019. Starting in 2020, the number and size of fires was randomly drawn from one of these historical year-sets for each timestep and Monte Carlo realization, using burn severity probabilities from @selmants_baseline_2017. Wildfire in the low land-use scenario was sampled from the subset of historical fire years at or below the median area burned statewide from 1999-2019. The high land-use scenario sampled from historical fire years above the median area burned over the same 21-year period (Fig. S2a).

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (figure S1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

@@

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in total ecosystem carbon storage (a) and individual carbon stocks (b) for the seven main Hawaiian Islands. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations."}
knitr::include_graphics("./fig_images/fig2_Cstocks.png")
```

Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land-use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (figure S4) and sustained losses of forest and shrubland land area in the high land-use scenario (figure S5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (figure S4) resulted in R~h~ being ~ 3% higher by 2100 than under RCP 4.5 (figure 3). Heterotrophic respiration declined substantially over time in the high land-use scenario (figure 3) because of long-term reductions in forest and shrubland land area (figure S5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land-use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land-use scenario and ~0.2 Tg y^-1^ in the low land-use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land-use scenario, driven primarily by greater variability in wildland fire probabilities.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in net primary production (NPP), heterotrophic respiration (Rh) and carbon fluxes induced by land use and land-use change for the seven main Hawaiian Islands. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations."}
knitr::include_graphics("./fig_images/fig3_Cflows.png")
```

### *3.2 Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios (figure 4). On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). The high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reductions and local avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in net biome productivity (NBP) for the seven main Hawaiian Islands. Values above zero indicate terrestrial ecosystems are acting as a net carbon sink for atmospheric carbon and values below zero indicate ecosystems are acting as a net carbon source to the atmosphere. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations. The dashed horizontal line in each panel represents the boundary between ecosystems acting as a net carbon sink (positive NBP values) and a net carbon source (negative NBP values)."}
knitr::include_graphics("./fig_images/fig4_NBPpanels.png") 
```

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in cumulative net biome productivity (NBP) for the seven main Hawaiian Islands when switching from the high to low land-use change scenario under each radiative forcing scenario (top panel) and when switching from the high (RCP 8.5) to low (RCP 4.5) radiative forcing scenario under each land-use scenario (bottom panel). Box plots indicate the median (vertical black line), 25th and 75th percentiles (colored boxes), 10th and 90th percentiles (thin horizontal lines), and values outside of this range (black circles). Note the different x-axis scales in each panel."}
knitr::include_graphics("./fig_images/fig5_cumdNBP.png") 
```

### *3.3 CO~2~ fertilization effect*
Projected estimates of both total ecosystem carbon storage and ecosystem carbon balance were highly sensitive to differing rates of a CFE on plant productivity. Under the high radiative forcing (RCP 8.5) and high land-use scenario, the inclusion of a CFE ranging from 5-15% led to ~33-98 Tg of additional carbon storage in ecosystems by the end of the century, a ~10-30% increase (figure 6a). Compared to the reference scenario (0% CFE), a 5% CFE was sufficient to transform Hawaiian Island ecosystems from a net carbon source to the atmosphere during the latter half of the 21st century (figure 4b) to a net carbon sink for the entire simulation period (figure 6b), completely offsetting all other carbon losses induced by high radiative forcing and high land use. Net carbon sink strength was further enhanced at higher CFE rates, with NBP increasing by an average of 0.07 Tg C y^-1^ for each 1% increase in CFE (figure 6b). When compared to other scenarios, applying a 5% CFE to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 0.46 ± 0.3 Tg C y^-1^, roughly equivalent to mean annual NBP in the low radiative forcing and low land-use scenario with no CFE (0.52 ± 0.12). A 15% CFE applied to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 1.18 ± 0.29 Tg C y^-1^, more than double that of the low radiative forcing and low land-use scenario with no CFE.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Sensitivity of projected changes in total ecosystem carbon storage (a) and mean annual net biome productivity (b) to different rates of carbon dioxide fertilization in the seven main Hawaiian Islands under the RCP 8.5 radiative forcing and high land use scenario. The carbon dioxide fertilization effect (CFE) is the percent change in net primary productivity (NPP) for every 100 ppm increase in atmospheric carbon dioxide. The CFE for all rates is capped at 600 ppm, which is achieved around the year 2060. Solid lines in (a) indicate the mean total ecosystem carbon storage across 30 Monte Carlo realizations for each CFE rate, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations. Solid circles in (b) represent mean annual net biome productivity averaged across all years and Monte Carlo realizations for each CFE rate, with vertical lines indicating the standard deviation of the mean. The dashed horizontal line in (b) represents the boundary between ecosystems acting as a net carbon sink (positive NBP values) and a net carbon source (negative NBP values)."}
knitr::include_graphics("./fig_images/fig6_cfe.png") 
```

## 4. Discussion

@@

### *4.1 Drivers of ecosystem carbon balance*

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2019 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019].

@@

## Data and Code Availability

>>>>

The State of Hawaiʻi exemplifies the challenges associated with projecting the interactive effects of future climate and land use change on ecosystem carbon balance at a regional scale. Hawaiʻi was the first U.S. state to enact legislation committing to full carbon neutrality, requiring the state to account for and offset all of its greenhouse gas emissions by 2045 (State of Hawaiʻi Acts 15 and 16). This legislation emphasizes the mitigation potential of natural ecosystems as a key component to emissions reduction, necessitating baseline estimates and future projections of land carbon sink strength. However, Hawaiʻi's challenging terrain complicates these assessment efforts. The main Hawaiian Islands are a complex mosaic of natural and human-dominated landscapes overlain by steep climate gradients across relatively short distances (supplmental figure 3), with mean annual temperature ranging from ~4-24° C [@giambelluca_evapotranspiration_2014] and mean annual rainfall ranging from ~180-9500 mm [@giambelluca_online_2013]. Temperatures have risen rapidly in the Hawaiian Islands since the mid 1970s [@giambelluca_secular_2008] and a long-term drying trend has persisted since the early 1920s [@frazier_spatial_2017], resulting in reduced forest biomass and productivity [@barbosa_effects_2017]. These same drying and warming trends have increased the frequency and intensity of wildland fire [@trauernicht_contemporary_2015; @trauernicht_vegetationrainfall_2019] with predictable negative effects on ecosystem carbon balance [@selmants_baseline_2017]. Ecosystem carbon stocks across the main Hawaiian Islands have also been strongly influenced by the legacy of past land-use change [@osher_effect_2003; @asner_high-resolution_2011]. Thousands of hectares of land were deforested beginning in the late 19th century to clear land for sugar plantations and cattle pasture [@cuddihy_alteration_1990]. Since the mid-20th century, much of this agricultural land has been steadily converted to urban areas, commercial forestry plantations, or simply abandoned and colonized by non-native grass species [@suryanata_diversified_2009; @perroy_evolving_2016]. Although these past trends surely inform the future impact of climate and land-use change on ecosystem carbon balance, high spatial and temporal heterogeneity complicates realistic projection efforts. To date only one study has attempted to integrate land-use, climate, and natural disturbances into future projections of Hawaiian ecosystem carbon balance, with projections limited to the mid-21st century under a single land-use change scenario and moderate radiative forcing [SRES A1B, equivalent to RCP 6; @selmants_baseline_2017].

@@

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CFE on NPP. The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### *2.1 Study area*
The spatial extent of this study was the terrestrial portion of the seven main Hawaiian Islands (figure 1), a total land area of 16,554 km^2^. We subdivided this landscape into a grid of 264,870 simulation cells, each of which was 250 x 250 m in size. Each simulation cell was assigned to one of 210 possible state types based on the unique combination of three moisture zones (dry, mesic, and wet; supplemental figure 1), seven islands, and ten discrete land cover classes (figure 1).

```{r fig.align = 'center', out.width = "99%", fig.cap = "Land cover classification of the seven main Hawaiian Islands, adapted from Jacobi \\textit{et al.} (2017). Agriculture in this map combines herbaceous and woody crops, but these two crop types are treated as separate land cover classes in the simulation model. Water and Wetland land cover classes are not shown."}
knitr::include_graphics("./fig_images/fig1_HI_LC.png")
```

### *2.2 States and transitions*
We developed two land-use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transitions between state types were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, shrubland, and tree plantation state types each had multiple transition pathways, while the barren state type could only transition to developed (i.e., urbanization). Although most state types had an urbanization transition pathway, there was no transition pathway out of an urbanized (developed) state. Water and wetland state types remained static throughout the simulation period.

@@

The wildfire transition sub-model was modified from @daniel_state-and-transition_2016 by incorporating a new 21-year historical wildfire spatial database of the Hawaiian Islands (supplemental figure 2). We used this new spatial database to calculate historical wildfire size distribution and ignition probabilities for each unique combination of moisture zone (supplemental figure 1), island, and state type (figure 1) for the years 1999-2019. Starting in 2020, the number and size of fires was randomly drawn from one of these historical year-sets for each timestep and Monte Carlo realization, using burn severity probabilities from @selmants_baseline_2017. Wildfire in the low land-use scenario was sampled from the subset of historical fire years at or below the median area burned statewide from 1999-2019. The high land-use scenario sampled from historical fire years above the median area burned over the same 21-year period (supplemental figure 2a).

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (supplmental figure 1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawaiʻi-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

@@

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in total ecosystem carbon storage (a) and individual carbon stocks (b) for the seven main Hawaiian Islands. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations."}
knitr::include_graphics("./fig_images/fig2_Cstocks.png")
```
Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land-use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (supplemental figure 4) and sustained losses of forest and shrubland land area in the high land-use scenario (supplmental figure 5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (supplmental figure 4) resulted in R~h~ being ~ 3% higher by 2100 than under RCP 4.5 (figure 3). Heterotrophic respiration declined substantially over time in the high land-use scenario (figure 3) because of long-term reductions in forest and shrubland land area (supplemental figure 5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land-use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land-use scenario and ~0.2 Tg y^-1^ in the low land-use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land-use scenario, driven primarily by greater variability in wildland fire probabilities.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in net primary production (NPP), heterotrophic respiration (Rh) and carbon fluxes induced by land use and land-use change for the seven main Hawaiian Islands. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations."}
knitr::include_graphics("./fig_images/fig3_Cflows.png")
```
### *3.2 Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios (figure 4). On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). The high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reductions and local avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in net biome productivity (NBP) for the seven main Hawaiian Islands. Values above zero indicate terrestrial ecosystems are acting as a net carbon sink for atmospheric carbon and values below zero indicate ecosystems are acting as a net carbon source to the atmosphere. Solid lines indicate the mean of 30 Monte Carlo realizations for each scenario, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations. The dashed horizontal line in each panel represents the boundary between ecosystems acting as a net carbon sink (positive NBP values) and a net carbon source (negative NBP values)."}
knitr::include_graphics("./fig_images/fig4_NBPpanels.png") 
```
```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected changes in cumulative net biome productivity (NBP) for the seven main Hawaiian Islands when switching from the high to low land-use change scenario under each radiative forcing scenario (top panel) and when switching from the high (RCP 8.5) to low (RCP 4.5) radiative forcing scenario under each land-use scenario (bottom panel). Box plots indicate the median (vertical black line), 25th and 75th percentiles (colored boxes), 10th and 90th percentiles (thin horizontal lines), and values outside of this range (black circles). Note the different x-axis scales in each panel."}
knitr::include_graphics("./fig_images/fig5_cumdNBP.png") 
```
### *3.3 CO~2~ fertilization effect*
Projected estimates of both total ecosystem carbon storage and ecosystem carbon balance were highly sensitive to differing rates of a CFE on plant productivity. Under the high radiative forcing (RCP 8.5) and high land-use scenario, the inclusion of a CFE ranging from 5-15% led to ~33-98 Tg of additional carbon storage in ecosystems by the end of the century, a ~10-30% increase (figure 6a). Compared to the reference scenario (0% CFE), a 5% CFE was sufficient to transform Hawaiian Island ecosystems from a net carbon source to the atmosphere during the latter half of the 21st century (figure 4b) to a net carbon sink for the entire simulation period (figure 6b), completely offsetting all other carbon losses induced by high radiative forcing and high land use. Net carbon sink strength was further enhanced at higher CFE rates, with NBP increasing by an average of 0.07 Tg C y^-1^ for each 1% increase in CFE (figure 6b). When compared to other scenarios, applying a 5% CFE to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 0.46 ± 0.3 Tg C y^-1^, roughly equivalent to mean annual NBP in the low radiative forcing and low land-use scenario with no CFE (0.52 ± 0.12). A 15% CFE applied to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 1.18 ± 0.29 Tg C y^-1^, more than double that of the low radiative forcing and low land-use scenario with no CFE.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Sensitivity of projected changes in total ecosystem carbon storage (a) and mean annual net biome productivity (b) to different rates of carbon dioxide fertilization in the seven main Hawaiian Islands under the RCP 8.5 radiative forcing and high land use scenario. The carbon dioxide fertilization effect (CFE) is the percent change in net primary productivity (NPP) for every 100 ppm increase in atmospheric carbon dioxide. The CFE for all rates is capped at 600 ppm, which is achieved around the year 2060. Solid lines in (a) indicate the mean total ecosystem carbon storage across 30 Monte Carlo realizations for each CFE rate, with shaded areas indicating the minimum and maximum range of Monte Carlo realizations. Solid circles in (b) represent mean annual net biome productivity averaged across all years and Monte Carlo realizations for each CFE rate, with vertical lines indicating the standard deviation of the mean. The dashed horizontal line in (b) represents the boundary between ecosystems acting as a net carbon sink (positive NBP values) and a net carbon source (negative NBP values)."}
knitr::include_graphics("./fig_images/fig6_cfe.png") 
```
## 4. Discussion

@@

### *4.1 Impact of different climate and land-use pathways* 
By comparing model results of Hawaiian ecosystem carbon balance under different scenario combinations, we were able to assess the relative impact of both global emissions reductions and regional actions to reduce emissions from land use, land use change, and disturbances such as wildland fire (figure 5). Global adherence to a lower emissions trajectory (i.e., switching from RCP 8.5 to RCP 4.5) had the largest cumulative impact, resulting in a median of 26 additional Tg C sequestered by Hawaiian ecosystems over the 90-year simulation period. Long-term reductions in the intensity of land-use change also consistently led to an increase in ecosystem carbon sequestration, but to a lesser degree than global emissions reductions. Switching from the ‘high’ to the ‘low’ land‐use scenario resulted in a median cumulative retention of an additional 11.6 Tg C in Hawaiian ecosystems by 2100. The combination of global climate mitigation and a reduction in land‐use conversions had the largest potential benefit to ecosystem carbon sequestration, reducing cumulative net losses by over 400% (37.7 Tg C). Notably,

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period {0.314 Tg C y^-1^; @selmants_baseline_2017]. Our current model estimates of NPP and R~h~ being 9.5% and 15% lower, respectively [

@@

## Data and code availability

### 2021-02-19 Manuscript/HI_Model_supp.Rmd
updated manuscript files

```{r fig.align = 'center', out.width = "99%", fig.cap = "Moisture zones of the seven main Hawaiian Islands, adapted from Jacobi et al (2017)."}
knitr::include_graphics("./fig_images/figS1_MZ.png")
```

## Wildland Fire

This study used a new spatial database of wildland fire perimeters in the Hawaiian Islands for the years 1999-2019 compiled using data from the Monitoring Trends in Burn Severity (MTBS) program, the Hawaiʻi Wildfire Management Organization, the U.S. National Park Service, the Oʻahu Army Natural Resources Program, and independent remote sensing analyses using publicly available LandSat data. This new wildland fire spatial database was used to calculate annual area burned (Figure S2), wildland fire probabilities, and wildland fire size distributions.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Annual area burned by wildland fire in the State of Hawaiʻi from 1999-2019, summed by year across the seven main Hawaiian Islands (A) and within each of the four largest islands (B). The dashed horizontal line in (A) represents the median area burned from 1999-2019."}
knitr::include_graphics("./fig_images/figS2_AnnualAreaBurned.png")
```

## Climate

Spatially explicit mean annual data on contemporary temperature and rainfall for the main Hawaiian Islands at 250-m resolution (Figure S3) are from @giambelluca_online_2013 and @giambelluca_evapotranspiration_2014. Spatially explicit projections of changes in annual temperature and rainfall by the year 2100 were developed using statistically downscaled CMIP5 climate projections for the Hawaiian Islands from @timm_statistical_2015 and @timm_future_2017 under RCPs 4.5 and 8.5 (Figure S5).

```{r fig.align = 'center', out.width = "99%", fig.cap = "Mean annual temperature (top panel) and mean annual rainfall (bottom panel) for the seven main Hawaiian Islands."}
knitr::include_graphics("./fig_images/figS3_CurrentClimate.png")
```

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected change in mean annual temperature (top panels) and mean annual rainfall (bottom panels) by end of century under RCPs 4.5 and 8.5 (top panels) based on statistical downscaling of an ensemble of climate models (CMIP5)."}
knitr::include_graphics("./fig_images/figS4_FutureClimate.png")
```

## Change in State Class Area

Projections of total land area covered by each State Class were summed by year across the seven main Hawaiian Islands from LUCAS model output data.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projections of total land area by year for each of five State Classes in the seven main Hawaiian Islands (Agriculture, Developed, Forest, Grassland, and Shrubland) under low and high land use change scenarios for the period 2010-2100. Solid lines represent the mean of 30 Monte Carlo realizations and shaded areas represent minimum and maximum Monte Carlo values."}
knitr::include_graphics("./fig_images/figS5_StateClassArea.png")
```

## References

>>>>

```{r fig.align = 'center', out.width = "99%", fig.cap = "Moisture zones of the seven main Hawaiian Islands, adapted from Jacobi \\textit{et al.} (2017)."}
knitr::include_graphics("./fig_images/figS1_MZ.png")
```
## Wildland Fire

This study used a new spatial database of wildland fire perimeters in the Hawaiian Islands for the years 1999-2019 using data compiled from the Monitoring Trends in Burn Severity (MTBS) program, the Hawaiʻi Wildfire Management Organization, the U.S. National Park Service, the Oʻahu Army Natural Resources Program, and independent remote sensing analyses using publicly available LandSat data. This new wildland fire spatial database was used to calculate annual area burned (Figure S2), wildland fire probabilities by Moisture Zone and State Class, and wildland fire size distributions.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Annual area burned by wildland fire in the State of Hawaiʻi from 1999-2019, summed by year across the seven main Hawaiian Islands (A) and within each of the four largest islands (B). The dashed horizontal line in (A) represents the median area burned from 1999-2019."}
knitr::include_graphics("./fig_images/figS2_AnnualAreaBurned.png")
```
## Climate

Spatially explicit contemporary mean annual temperature and rainfall data for the main Hawaiian Islands at 250-m resolution (Figure S3) are from @giambelluca_online_2013 and @giambelluca_evapotranspiration_2014. Spatially explicit projections of changes in annual temperature and rainfall by the year 2100 were developed using statistically downscaled CMIP5 climate projections for the Hawaiian Islands from @timm_statistical_2015 and @timm_future_2017 under RCPs 4.5 and 8.5 (Figure S5).

```{r fig.align = 'center', out.width = "95%", fig.cap = "Mean annual temperature (top panel) and mean annual rainfall (bottom panel) for the seven main Hawaiian Islands. Data from Giambelluca \\textit{et al.} (2013) and Giambelluca \\textit{et al.} (2014)."}
knitr::include_graphics("./fig_images/figS3_CurrentClimate.png")
```
```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected change in mean annual temperature (top panels) and mean annual rainfall (bottom panels) by end of century under RCPs 4.5 and 8.5 (top panels) based on statistical downscaling of an ensemble of climate models (CMIP5)."}
knitr::include_graphics("./fig_images/figS4_FutureClimate.png")
```
## Change in State Class Area

Projections of total land area covered by each State Class were summed by year and Monte Carlo iteration across the seven main Hawaiian Islands from LUCAS model output data.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projections of total land area by year for each of five State Classes in the seven main Hawaiian Islands (Agriculture, Developed, Forest, Grassland, and Shrubland) under low and high land use change scenarios for the period 2010-2100. Solid lines represent the mean of 30 Monte Carlo realizations and shaded areas represent minimum and maximum Monte Carlo values."}
knitr::include_graphics("./fig_images/figS5_StateClassArea.png")
```
## References
